Testing for measurement invariance with respect to an ordinal variable

### 2012-08-22 ordinal/ordinal.Rnw
Revised and added to the general discussion of the ordinal paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@165 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{enumerate}
  \item To my knowledge, these are the only proposed statistics that
    treat ordinal auxiliary 
    variables as truly ordinal.  LRT treats them as categorical,
    inclusion in the model typically treats tham as continuous.
  \item The first point is important because, if we test w.r.t.\ an
    ordinal auxiliary variable, we are generally looking for ordinal
    violations.  We don't care if a middle level differs from
    all others because we would not usually have an intuitive
    explanation for it.
  \item If we know our sample size is large and have an ordinal
    auxiliary variable, should we even use the LRT in
    the first place?  If the measurement invariance violation is truly
    ordinal, the proposed statistics should be better.  If it is not
    ordinal, the violation will be difficult to explain as anything
    but an anomaly.  The stakes might also play a role here.  In
    high-stakes testing, you might be more interested in treating the
    ordinal variable as categorical to detect any possible
    instability, even those that don't make sense.
  \item The proposed statistics should generally transfer to IRT and
    other psychometric models.
\end{enumerate}

\subsection{Summary}

>>>>

In this paper, we proposed two statistics for studying measurement
invariance that can be used when one has an auxiliary ordinal
variable.    To our knowledge,
the ordinal measurement invariance statistics proposed here are the
only ones that treat auxiliary variables as ordinal.  Other methods
treat the auxiliary variable as either continuous or categorical, in a
manner similar to the treatment of ordinal predictor variables in
linear regression.  In the remainder of the paper, we provide 
detail on test choice and on the tests' applicability to other
models.

\paragraph{Choice of Test}
The results presented in this paper imply that the proposed
ordinal statistics may ``miss'' measurement invariance violations that
are not monotonic w.r.t.\ $V$, moreso than the likelihood ratio test.  We
speculate that, in most applications, this will not be a major issue
because the researcher's a priori hypotheses exclusively focus on
monotonic measurement 
invariance violations.  For example, in the youth gratitude
application, we tested for measurement invariance across six age
groups.  If we observed a measurement invariance violation whereby
factor loadings were equal at all age groups except 14 years,
we would
have a hard time explaining the violation as anything but an anomaly
in the 14-year-olds.  Further, if $n$ is large, we are likely to
suspect that the result arises from the large sample size.
We may still be interested in why the
14-year-olds differed, but the analysis is purely exploratory at this
point because this type of violation was unexpected.  
However, there is generally a tradeoff between the ordinal statistics
and the likelihood ratio statistic.  The
proposed ordinal statistics usually provide sharper tests of one's a
priori hypothesis regarding measurement invariance w.r.t.\ ordinal
$V$, while the likelihood ratio statistic provides a sharper test of
general (non-monotonic) measurement invariance w.r.t.\ $V$.  While the
latter feature may be important in some high-stakes applications, many
researchers are likely to find the former feature appealing for their
work.

\paragraph{Extension to Other Models}
We focused on testing for measurement invariance in factor
analysis models here, but the proposed test statistics are applicable to
other psychometric models that are estimated via ML.  The only
requirement for carrying out the tests is that the casewise scores
(Equation~\eqref{eq:score}) be available following model estimation.
As a result, applications to studying DIF in IRT are immediate,
as are general psychometric applications to studying parameter
stability w.r.t.\ ordinal auxiliary variables.  We expect the
same general results to hold for these applications, whereby the
proposed test statistics are better than the LRT for detecting
monotonic instabilities.   The strucchange package \cite{Zei06} noted
previously can be used for these more-general applications.

\subsection{Summary}
The test statistics proposed in this paper have high power for
detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Bent80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous 
(e.g., \citeNP{BauHus09,MolDol10,NeaAgg06}).  In all, the tests have
advantageous properties that should be useful in practice.
